Challenges and Solutions of Developing and Implementing a Revolutionary Novel Desktop-as-a-Service

### 2024-04-26 paper.tex
Many language improvemens

Stability in Desktop as a Service (DaaS) applications is of major importance
to ensure that our solution can meet the demands of continuous
and time-critical operations across varied user scenarios.
From our point of view any daas system must ensure the stability
of all system components during runtime
in order to maintain the ability to provide service continuity,
handle growing user loads and to transparently recover from failures,
while still providing a seamless end-user experience.
Such considerations are an integral part of our architectural concept
as they directly impact the operational integrity
of the system and its hosted instances,
making them suitable for enterprise-level deployment and trust.

This definition of stability is underpinned by several architectural
and operational strategies including redundancy of components, scalability of nodes,
and a robust backend service framework
providing means of communication between components and nodes.
These strategies as well as their drawbacks and benefits
will be further discussed on a component level
and within the following subsections.

\subsubsection{Scalability of Components}
Scalability in general, but also specificaly in daas environments
can be approached in two fundamental ways:
horizontal and vertical scaling.
Horizontal scaling, or scaling out, involves adding more machines or instances
to a pool to manage increased load.
In contrast to that, vertical scaling or scaling up,
refers to adding more power (CPU, RAM)
to an existing machine~\cite{vaquero2011dynamically}.

@@

In order to comply to both we defined means of configuration
such that the system itself as well as all of its backend services
are designed to be split up into logical units
which can be distributed to dedicated systems.
Esssentially, all backend services receive their own dedicated configuration
and can be accessed via tuples of hostname and port.
Any component using such a backend service can obtain the relevant tuple
and then use it to fetch requests from that service.
By default and in a standlone configuration
all backend-services reside on the same native host
and relevant tuples can be derived from the host configuration.
In a distributed context a selection of these services
can be distributed to a dedicated and remote system
in order to comply to horizontal scaling principles.
The relevant tuple can then be derived from any node in the system
by utilizing a common database containing the relevant metainformation.

Vertical Scaling of daas nodes is currently not planned
but a adminstrators are free to use custom 3rd party toolchains, e.g for GPU virtualization.
Allthough vertical scaling on a host basis
is beyond scope of our research project,
it is more plausbile to scale the instances
being hosted within that system and for both dimensions.

For vertical scaling of instances appropriate parameters and api endpoints are present
in order to manipulate the configuration of their resource assignments,
such that arbitrary configurations are possible.

For horizontal scaling of instances
the builtin mechanisms of each backend service are automatically configured and utilized
such that instances can be arbitrily distributed across a pool of connected nodes.
In the case of virtual machines the responsible backend service proxmox implements
such mechanisms as part of its 'Proxmox VE High Availability (HA)' cluster.
In the case of containers any backend service fullfilling
similar concepts can be integrated, as for example with kubernetes.
Within such clusters any daas instance is fully agnostic to its hosting system
and must only be reachable via a known hostname or ip.

@@

Redundancy is critical for most distributed systems
but especially for daas architectures,
and serves as a safeguard against service interruptions and data loss
by duplicating essential components, functions and data.
This design principle ensures that systems remain highly available and reliable,
even during component failures.
In the following a list of components eligable for redundancy is given
which briefly describes the most relevant strategies contained in our architecture.

\begin{itemize}
	\item \textbf{Frontend/Backend: } In our final version
	      redundancy of the frontend and backend components will offer means of configuration
	      such that multiple instances can be distributed to dedicated systems
	      and can be target of standardized load balancing strategies.
	      By doing so the frontend or the backend may just act as a gateway
	      between the user and the backend service containing the relevant instances.
	      In that manner redundancy is achieved and frontend or backend
	      can forward requests and responses
	      by identifying the correct communication endpoint
	      based on url's or specific parts of them,
	      as for example by extracting hostnames
	      or specific identifiers being unique within the daas domain.
	\item \textbf{Database: } The database is integrated as a backend service
	      and as such completely unaware of surrounding components
	      and contexts in which it is used.
	      The database can therefore be distributed to a dedicated system if needed.
	      Generally, any modern database (e.g. MariaDB) implements rich features
	      to maintain specific data consistency strategies
	      such as data replication, clustering,
	      dedicated failover mechnisms or also RAID control.
	      Therefore redundancies of our database system are currently not planned,
	      but can easily be applied by users with such needs
	      and by using database specific features and a modified database adapter.
	\item \textbf{Authentication: } Authentication within our architecture
	      is also implemented as a backend service
	      and can therefore be distributed to any dedicated system,
	      can be a target of standardized load balancing mechanisms and strategies
	      and is as well fully agnostic of surrounding components
	      and calling contexts.
	      The authentication system maintains
	      its own credential registry in a seperate database,
	      a customizable set of user and group privileges,
	      and implements the well defined OAuth 2.0 standard
	      to allow remote authentication.
	      In our final version, a customizable amount of redundant systems might be configured.
	      % TODO: Discuss: Is that true? Do we want to make Darios authenticator redundant?
	\item \textbf{Distributed Filesystem: } Within our architecture
	      we use the distributed filesystem ceph which is a distributed storage system
	      providing object, block and file storage. In ceph redundancy is achieved
	      through its self-healing and self-managing capabilities utilizing an algorithm
	      called CRUSH (Controlled Replication Under Scalable Hashing).
	      Data availability and fault-tolerance is obtained
	      by automatically managing replication
	      of data pieces across different devices.
	      Our designed architecture uses these capabilities
	      to reliably provide
	      file storage to host systems and all hosted daas instances
	      and provides block storage to the infrastructural backend services.
	      By hosting daas instances directly from the distributed block storage,
	      arbitrary redundancy scenarios might be covered
	      by using specific configuration parameters.
	\item \textbf{Instances: } Redundancy of instances
	      is not a necessary condition for daas systems in general
	      but it might make sense to apply such strategies under certain conditions.
	      By using highly available clusters within our backend services
	      redundancy can simply be achieved by using the corresponding api endpoints
	      previously described and by parameterizing them appropriately.
\end{itemize}

\subsubsection{Internal Communication}
After discussing the principles being used to achieve scalability and redundancy
it adheres to describe how communication flows are maintained
in such a distributed enironment.

Our proposed solution is at the core designed
as a message oriented architecture
which additionally aims to fullfill daas properties
for scalability and redundancy as previously described.

That implies fundamentally, that any action being triggered,
may it either be in the frontend on the client side,
in the backend on the host side
or within any of the hosted instances
must always fullfill the condition that the required information to execute that action
must either be directly derivable from a request or a corresponding response
or otherwise, must be immidiately available to the component executing that action.

@@

By doing so, any change being trigerred by a node within the daas network
is immidiately available to any other node in the system
which relies on that information change.
Apart from that any additional information needed by a component
is fully contained within the related request or response.

Therefore, arbitrary distributed configurations are possible
as long as request and response messages reach their desired target eventually
and as long availability of all relevant backend services is guaranteed.

If that is the case, each distributed daas node can typically
just contain all time-critical components as a bare minimum
in order to maintain an acceptable performance throughout the whole user-facing network stack.
This would typically include the backend service hosting the relevant instances
as well as the websocket proxy directly connected to the guacamole daemon
and all other components may reside somewhere in the daas network.

>>>>

Stability in Desktop as a Service (DaaS) applications is of significant importance
to ensure that our solution can meet the demands of continuous
and time-critical operations across varied user scenarios.
From our point of view, any DaaS system must ensure the stability
of all system components during runtime
to maintain the ability to provide service continuity,
handle growing user loads and transparently recover from failures,
while still providing a seamless end-user experience.
Such considerations are an integral part of our architectural concept
as they directly impact the operational integrity
of the system and its hosted instances,
making them suitable for enterprise-level deployment and trust.

This definition of stability is underpinned by several architectural
and operational strategies, including redundancy of components, scalability of nodes,
and a robust backend service framework
providing means of communication between components and nodes.
These strategies, as well as their drawbacks and benefits
will be further discussed on a component level
and within the following subsections.

\subsubsection{Scalability of Components}

Scalability in general, but also specifically in DaaS environments
can be approached in two fundamental ways:
Horizontal and vertical scaling.
Horizontal scaling, or scaling out, involves adding more machines or instances
to a pool to manage the increased load.
In contrast to that, vertical scaling or scaling up,
refers to adding more power (CPU, RAM)
to an existing machine~\cite{vaquero2011dynamically}.

@@

To comply with both, we defined means of configuration
such that the system itself, as well as all of its backend services
are designed to be split up into logical units
which can be distributed to dedicated systems.
Essentially, all backend services receive their dedicated configuration
and can be accessed via tuples of hostname and port.
Any component using such a backend service can obtain the relevant tuple
and then use it to fetch requests from that service.
By default and in a standalone configuration,
all backend services reside on the same native host, and relevant tuples can be derived from the host configuration.
In a distributed context, a selection of these services
can be distributed to a dedicated and remote system
to comply with horizontal scaling principles.
The relevant tuple can then be derived from any node in the system
by utilizing a common database containing the relevant meta information.

Vertical scaling of DaaS nodes is currently not planned
but administrators are free to use custom 3rd party toolchains, e.g., for GPU virtualization.
Although vertical scaling on a host basis
is beyond the scope of our research project,
it is more plausible to scale the instances
being hosted within that system and for both dimensions.

For vertical scaling of instances, appropriate parameters and API endpoints are present
to manipulate the configuration of their resource assignments,
such that arbitrary configurations are possible.

For horizontal scaling of instances
the builtin mechanisms of each backend service are automatically configured and utilized
such that instances can be arbitrarily distributed across a pool of connected nodes.
In the case of virtual machines, the responsible backend service proxmox implements
such mechanisms as part of its 'Proxmox VE High Availability (HA)' cluster.
In the case of containers, any backend service fulfilling
similar concepts can be integrated, for example, with Kubernetes.
Within such clusters, any DaaS instance is entirely agnostic to its hosting system
and must only be reachable via a known hostname or IP.

@@

Redundancy is critical for most distributed systems
but especially for DaaS architectures,
and serves as a safeguard against service interruptions and data loss
by duplicating essential components, functions, and data.
This design principle ensures that systems remain highly available and reliable,
even during component failures.
In the following, a list of components eligible for redundancy is given
which briefly describes the most relevant strategies contained in our architecture.

\begin{itemize}
\item \textbf{Frontend/Backend:} In our final version
redundancy of the frontend and backend components will offer means of configuration
such that multiple instances can be distributed to dedicated systems
and can be the target of standardized load-balancing strategies.
By doing so, the frontend or the backend may act as a gateway
between the user and the backend service containing the relevant instances.
In that manner, redundancy is achieved, and frontend or backend
can forward requests and responses
by identifying the correct communication endpoint
based on URL's or specific parts of them,
for example, by extracting hostnames
or specific identifiers being unique within the DaaS domain.
\item \textbf{Database:} The database is integrated as a backend service
and as such, completely unaware of surrounding components
and contexts in which it is used.
The database can, therefore, be distributed to a dedicated system if needed.
Generally, any modern database solution (e.g., MariaDB) implements rich features
to maintain specific data consistency strategies
such as data replication, clustering,
dedicated failover mechanisms or also RAID control.
Therefore, redundancies in our database system are currently not planned,
but can easily be applied by users with such needs
and by using database-specific features and a modified database adapter.
\item \textbf{Authentication: } Authentication within our architecture
is also implemented as a backend service
and can, therefore, be distributed to any dedicated system,
can be a target of standardized load-balancing mechanisms and strategies
and is as well fully agnostic of surrounding components
and calling contexts.
The authentication system maintains
its credential registry in a separate database,
a customizable set of user and group privileges,
and implements the well-defined OAuth 2.0 standard
to allow remote authentication.
A customizable amount of redundant systems might be configured in our final version.
	      % TODO: Discuss: Is that true? Do we want to make Darios authenticator redundant?
\item \textbf{Distributed Filesystem:} Within our architecture
we use the distributed filesystem Ceph, which is a distributed storage system
providing object, block, and file storage. In Ceph, redundancy is achieved
through its self-healing and self-managing capabilities utilizing an algorithm
called CRUSH (Controlled Replication Under Scalable Hashing).
Data availability and fault tolerance are obtained
by automatically managing replication
of data pieces across different devices.
Our designed architecture uses these capabilities
to reliably provide
file storage to host systems and all hosted DaaS instances
and provides block storage to the infrastructural backend services.
By hosting DaaS instances directly from the distributed block storage,
arbitrary redundancy scenarios might be covered
by using specific configuration parameters.
\item \textbf{Instances:} Redundancy of instances
It is not a necessary condition for DaaS systems in general, but applying such strategies under certain conditions might make sense.
By using highly available clusters within our backend services
redundancy can be achieved by using the corresponding API endpoints
previously described and by parameterizing them appropriately.
\end{itemize}

\subsubsection{Internal Communication}
After discussing the principles being used to achieve scalability and redundancy
it adheres to describe how communication flows are maintained
in such a distributed environment.

Our proposed solution is, at the core, designed
as a message-oriented architecture
which additionally aims to fulfill DaaS properties
for scalability and redundancy as previously described.

That implies fundamentally that any action being triggered,
may it either be in the frontend on the client side,
in the backend on the host side
or within any of the hosted instances
must always fulfill the condition that the required information to execute that action
must either be directly derivable from a request or a corresponding response
or otherwise, must be immediately available to the component executing that action.

@@

By doing so, any change being triggered by a node within the DaaS network
is immediately available to any other node in the system
which relies on that information change.
Apart from that, any additional information needed by a component
is fully contained within the related request or response.

Therefore, arbitrarily distributed configurations are possible
as long as request and response messages reach their desired target, eventually
and as long as the availability of all relevant backend services is guaranteed.

If that is the case, each distributed DaaS node can typically
just contain all time-critical components at a bare minimum
to maintain an acceptable performance throughout the whole user-facing network stack.
This would typically include the backend service hosting the relevant instances
as well as the WebSocket proxy directly connected to the guacamole daemon
and all other components may reside somewhere in the DaaS network.
